---
title: "FSR Plotting and Basic Stats"
author: "Daniel Wilkinson"
format: pdf
editor: visual
---

In a similar way to the plasma leucine analyses, we are interested in visualising the FSR data and performing some simple stats for each group.

```{r}
{r}
# load in libraries
library(tidyverse) # for data loading and manipulation
library(tidyplots) # for publication ready plots
library(lme4) # for linear model analyses
library(garfify) # for post-hoc constrasts
library(emmeans) # additional post-hoc options
```

First step is to create publication ready visulisations for each re-feeding group, we are interested here in looking at the groups individually and in the same plot for visual comparison.

```{r}
# load in data and plot - and adjust with additions and changes
early <- read_csv("fsr_early.csv")
mid <- read_csv("fsr_mid.csv")
late <- read_csv("fsr_late.csv")

# combine for single plot
fsr <- bind_rows(early, mid, late)
fsr <- fsr |> 
  mutate(group = fct_relevel(group,c("Early", "Mid", "Late")))


timecourseFSR <- tidyplot(fsr, x = time, y = fsr, color = group) |>
  add_mean_dot(size = 3) |>
  add_mean_line(linewidth = 1.25, alpha = 0.5) |>
  add_sem_errorbar(width = 10) |>
  theme_tidyplot() |>
  adjust_size(width = 150, height = 80) |>
  adjust_x_axis_title(title = "Time (mins)") |>
  adjust_y_axis_title("Myofibrillar FSR (%/h)") |>
  adjust_y_axis(limits = c(0,0.12)) |>
  adjust_font(fontsize = 12) |>
  adjust_legend_title(title = "Re-Feed Group") |>
  save_plot("combinedFSR.pdf")
```

Next up is a simple linear model statistical analyses with post hoc comparisons to determine where significant increases in FSR occur compared to baseline values in response to the feed and the re-feed periods for each group. First summaries and normal distribution tests are run.

```{r}
# summary stats and normality checks
# set as factors
early$time_period <- factor(early$time_period, levels = c("Basal", "0-90", "90-180", "180-270"))
mid$time_period <- factor(mid$time_period, levels = c("Basal", "0-90", "90-180", "180-240", "240-330"))
late$time_period <- factor(late$time_period, levels = c("Basal", "0-90", "90-180", "180-300", "300-390"))

# summarise data in tables
# early
summary <- early |>
  group_by(time_period) |>
  get_summary_stats(fsr, type = "mean_sd")

sum_early <- data.frame(summary)
write_csv(sum_early, "summaryFSRData_early.csv")

# assess for normal distribution
normality <- early |>
  group_by(time_period) |>
  shapiro_test(fsr)

norm_early <- data.frame(normality)
write_csv(norm_early, "normalityFSRData_early.csv")

# mid
summary <- mid |>
  group_by(time_period) |>
  get_summary_stats(fsr, type = "mean_sd")

sum_mid <- data.frame(summary)
write_csv(sum_mid, "summaryFSRData_mid.csv")

# assess for normal distribution
normality <- mid |>
  group_by(time_period) |>
  shapiro_test(fsr)

norm_mid <- data.frame(normality)
write_csv(norm_mid, "normalityFSRData_mid.csv")

# late
summary <- late |>
  group_by(time_period) |>
  get_summary_stats(fsr, type = "mean_sd")

sum_late <- data.frame(summary)
write_csv(sum_late, "summaryFSRData_late.csv")

# assess for normal distribution
normality <- late |>
  group_by(time_period) |>
  shapiro_test(fsr)

norm_late <- data.frame(normality)
write_csv(norm_late, "normalityFSRData_late.csv")

```

Some missing samples timepoints in the mid and late group due to sample processing issues, loss of sample in processing etc.. However all show normal distribution except for one group Mid 0 timepoint, however with such small n for each group, we can assume that this is inconsequential to the overall analyses. Now linear models are run to look for significant changes in FSR due tot he fedding in each group.

```{r}
# Random intercept and random slope for time per subject and simple linear model run using lme4 package with subjects as random variable
# early
model <- lmer(fsr ~ time_period +(1 | subject), data = early)

# View the ANOVA-style summary table - no main effect of time for early fsr, contrasts run anyway
anova_early <- anova(model)

## now look at contrasts for post-hoc analyses
contrastFSR_early <- emmeans(model, pairwise ~ time_period)
write_csv(data.frame(contrastFSR_early$contrasts), "lm_constrastsFSR_early.csv")

# boxplot of early group
fsrBoxplot_Early <- tidyplot(early, x = time_period, y = fsr) |>
  add_boxplot(box_width = 0.5, alpha = 0.5, whiskers_width = 0.5) |>
  add_data_points_jitter(size = 2, alpha = 0.5) |>
  theme_tidyplot() |>
  adjust_size(width = 150, height = 80) |>
  adjust_colors("#0072B2") |>
  adjust_x_axis_title(title = "FSR Time Period") |>
  adjust_y_axis_title("Myofibrillar FSR (%/h)") |>
  adjust_y_axis(limits = c(0,0.16)) |>
  adjust_font(fontsize = 12) |>
  save_plot("fsrBoxPlotEarly.pdf")

# mid
model <- lmer(fsr ~ time_period + (1 | subject), data = mid)

# View the ANOVA-style summary table - no main effect of time for early fsr, contrasts run anyway
anova_mid <- anova(model)

## now look at contrasts for post-hoc analyses
contrastFSR_mid <- emmeans(model, pairwise ~ time_period)
write_csv(data.frame(contrastFSR_mid$contrasts), "lm_constrastsFSR_mid.csv")

# boxplot of mid group
fsrBoxplot_Mid <- tidyplot(mid, x = time_period, y = fsr) |>
  add_boxplot(box_width = 0.5, alpha = 0.5, whiskers_width = 0.5) |>
  add_data_points_jitter(size = 2, alpha = 0.5) |>
  theme_tidyplot() |>
  adjust_size(width = 150, height = 80) |>
  adjust_colors("#009E73") |>
  adjust_x_axis_title(title = "FSR Time Period") |>
  adjust_y_axis_title("Myofibrillar FSR (%/h)") |>
  adjust_y_axis(limits = c(0,0.16)) |>
  adjust_font(fontsize = 12) |>
  save_plot("fsrBoxPlotMid.pdf")

# late
model <- lmer(fsr ~ time_period +(1 | subject), data = late)

# View the ANOVA-style summary table - no main effect of time for early fsr, contrasts run anyway
anova_late <- anova(model)

## now look at contrasts for post-hoc analyses
contrastFSR_late <- emmeans(model, pairwise ~ time_period)
write_csv(data.frame(contrastFSR_late$contrasts), "lm_constrastsFSR_late.csv")

# boxplot of late group
fsrBoxplot_Late <- tidyplot(late, x = time_period, y = fsr) |>
  add_boxplot(box_width = 0.5, alpha = 0.5, whiskers_width = 0.5) |>
  add_data_points_jitter(size = 2, alpha = 0.5) |>
  theme_tidyplot() |>
  adjust_size(width = 150, height = 80) |>
  adjust_colors("#D55E00") |>
  adjust_x_axis_title(title = "FSR Time Period") |>
  adjust_y_axis_title("Myofibrillar FSR (%/h)") |>
  adjust_y_axis(limits = c(0,0.16)) |>
  adjust_font(fontsize = 12) |>
  save_plot("fsrBoxPlotLate.pdf")

```

Simple linear model analyses complete, and outputs saved to file.
